---
title: "Principal Component Analysis using R"
author: "Joshua Edefo"
date: "2024-07-27"
email: "edefojoshua2000@yahoo.com"
output: github_document
---
library

```{r a, message = FALSE}
library(ggplot2)
```
Load the iris dataset

```{r b}
data(iris)
head(iris)
str(iris)
summary (iris)
```
Exclude the species column as it's a categorical variable

```{r c}
iris_data <- iris[, 1:4] # or iris_data <- iris[, -5] or iris_data <- iris (~Sepal.Length + Sepal.Length + Petal.Length + Petal.Width, data = iris )
iris_data
```

Standardised the scale
```{r d}
iris_scaled <- scale(iris_data)
```

Perform PCA, &  view eigen values and vectors
```{r e}
pca_result <- prcomp(iris_scaled, center = TRUE, scale. = TRUE)
pca_result
```
plot showing the variance $ visualize the explained variance (scree plot)

```{r g}
plot(pca_result, type = "l")
scree_plot <- qplot(1:length(pca_result$sdev), 
                    (pca_result$sdev)^2 / sum((pca_result$sdev)^2), 
                    geom = "line") +
  geom_point() +
  labs(x = "Principal Component", 
       y = "Proportion of Variance Explained",
       title = "Scree Plot")
print(scree_plot)
```

Visualize the PCA results (biplot), extract PC scores & extract and view the principal components

```{r z}
biplot(pca_result, scale = 0)
str(pca_result)
pca_result$x
```
Take PC scores of PCI and PC2 and attached to iris data

```{r i}
iris2 <- cbind (iris,pca_result$x [, 1:2])
head(iris2)
str (iris2)
```
Plot with ggplot

```{r j}
ggplot(iris2, aes(PC1, PC2, col = Species, fill = Species)) + 
  stat_ellipse(geom = "polygon", col = "black", alpha = 0.5) + 
  geom_point(shape = 21, col = "black") 
```

 Correlations between variables and Principal Component Scores
```{r k}
cor (iris[, -5])

cor (iris[, -5], iris2 [, 6:7] )

```

session information
```{r l}
sessionInfo()
```
